Self-assembly of colloid-cholesteric composites: A route to switchable optical materials

### 2013-08-30 papers/2012-sandwich/support.tex
Revised support - Davide, Aug 30
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@2142 40876441-3496-0410-a613-a2fca35b2ae2

{\bf Supporting Online Material: Sandwich Title}

Here, the details of the simulation method, protocol, and parameters are
set out. The matching of simulation units to physical units is also
discussed.

@@

The remaining terms of the free energy density in Eq.~\ref{free} describe
distortions of
the order parameter field. It is conventional \cite{blue1,deGennes} to
assume that splay, bend and twist deformations of the director are equally
costly, that is, there is a single elastic constant $K$. The parameter $q_0$
is related to the helical pitch length $p$ via $q_0=2\pi/p$,
describing one full turn of the director in the cholesteric phase.

@@

@@

The tensor $S_{\alpha\beta}$ in Eq.~\ref{eom} completes the material
derivative for rod-like molecules \cite{beris}. It couples the order
parameter to the symmetric and antisymmetric parts of the velocity 
gradient tensor $W_{\alpha \beta}\equiv\partial_\beta u_\alpha$.
The symmetric part $A_{\alpha\beta}$ and the antisymmetric part
$\Omega_{\alpha\beta}$ are defined as 
\begin{equation}
A_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} + W_{\beta\alpha}),
\end{equation}
and
\begin{equation}
\Omega_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} - W_{\beta\alpha}).
\end{equation}
This full coupling term is then
\begin{eqnarray}
\label{coupling-term}
S_{\alpha\beta}  &=&
(\xi A_{\alpha\nu} + \Omega_{\alpha\nu})
(Q_{\nu\beta} + {\textstyle \frac{1}{3}}\delta_{\nu\beta})
+
(Q_{\alpha\nu} + {\textstyle \frac{1}{3}} \delta_{\alpha\nu})
(\xi A_{\nu\beta} - \Omega_{\nu\beta})
\nonumber\\ 
&-&2 \xi ({Q_{\alpha\beta} + {\textstyle \frac{1}{3}}\delta_{\alpha\beta}})
Q_{\nu\nu} W_{\nu\nu}.
\end{eqnarray}
CORRECT TRACE?!
Here, $\xi$ is a material-dependent `tumbling parameter' that controls the
relative importance of rotational and elongational flow for molecular
alignment. In all cases, the value used here is $\xi = 0.7$, which is within
the `flow aligning' regime, where molecules align at a fixed angle
(the Leslie angle) to the flow direction in weak simple shear \cite{deGennes}.
The value of the collective rotational diffusion used in all simulations is
$\Gamma = 0.5$ in simulation units.

@@

In all the simulations reported here the preferred orientation of the
director field at the colloid surface is normal to the surface. The
required nematic director at the particle surface in a given location
$\hat{n}^0_\alpha$ may then be determined from geometry alone,
and a preferred order parameter $Q_{\alpha\beta}^0$ is computed via
\begin{equation}
Q^0_{\alpha\beta} = q^0(\hat{n}_\alpha^0 \hat{n}_\beta^0 
- {\textstyle \frac{1}{3}} \delta_{\alpha\beta}).
\label{eq:surf_q}
\end{equation}
The magnitude of surface order $q^0$ is set by
\begin{equation}
q^0 = {\textstyle \frac{2}{3}} \left( {\textstyle \frac{1}{4}} 
+ {\textstyle \frac{3}{4}} \sqrt{1 + (8/3\gamma)} \right)
\label{eq:surf_q0}
\end{equation}
CHECK which corresponds to SOMETHING [REFERENCE].

In all simulations the colloidal (hard sphere) radius is $a = 2.3$. To
prevent colloids overlapping at the hard sphere radius, an addition
short range soft potential is included. This takes the form of
$V(h) = \epsilon (\sigma/h)$ where $h$ is the surface to surface separation.
The parameters are $\epsilon = 0.0004$ and $\sigma = 0.1$ in simulation
units. Both potential and resulting force are smoothly matched to zero at
a cut-off distance of $h_c = 0.25$ simulation units.

@@

ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK

ISSUE: no redshift?

ISSUE: do we want some free energy numbers?

PENDING

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.35]{s1.jpg}
\end{center}
\caption{As for main manuscript Fig.~1 (initialised to equilibrium blue
phase I), but for the entire simulation
system of 128$^3$ lattice sites. (a) solid volume fraction 1\% and
weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
volume fraction and strong anchoring. The view direction in the
main figure is from the left here. There is a reference
particle at the bottom left in each panel which does not take part
in the simulation.}
\end{figure}

\begin{figure}
\includegraphics[scale=0.35]{s2.jpg}
\caption{As for main manuscript Fig.~2 (initialised via a ``quench''
to generate a disordered network), but for the entire simulation
system of 128$^3$ lattice sites. (a) solid volume fraction 1\% and
weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
volume fraction and strong anchoring. Again, there is a reference
particle at the bottom left in each case which does not take part
in the simulation.}
\end{figure}

@@

@@

>>>>

{\bf Supporting Information for: Self-assembly of colloid-cholesteric composites: A route to switchable optical materials}

Here, we give more details of the simulation method, protocol, and parameters used. We also provide additional figures complementing the main text, and discuss the matching of simulation units to physical units.

@@

The remaining terms of the free energy density in Eq.~\ref{free} describe
distortions of the order parameter field. In theoretical work and when
describing a generic rather than a specific system, it is 
conventional~\cite{blue1,deGennes} to
assume that splay, bend and twist deformations of the director are equally
costly, that is, there is a single elastic constant $K$. The parameter $q_0$
is related to the helical pitch length $p$ via $q_0=2\pi/p$,
describing one full turn of the director in the cholesteric phase.

@@

The effect of an electric field can be modelled by including the following
term in the free energy density,
\begin{equation}
-(\epsilon_a/12\pi)E_{\alpha}Q_{\alpha\beta}E_{\beta}
\end{equation} 
where $E_{\beta}$ are the components of the electric field, and 
$\epsilon_a$ (here assumed positive) is the dielectric anisotropy of the liquid
crystal. The strength of the electric field is quantified via
a further dimensionless number, 
\begin{equation}
{\cal E}^2=(27 \varepsilon_a E_\alpha E_\alpha)/(32 \pi A_0 \gamma).
\end{equation}

@@

The tensor $S_{\alpha\beta}$ in Eq.~\ref{eom} completes the material
derivative for rod-like molecules \cite{beris}. It couples the order
parameter to the symmetric and antisymmetric parts of the velocity 
gradient tensor $W_{\alpha \beta}\equiv\partial_\beta u_\alpha$.
The symmetric part $A_{\alpha\beta}$ and the antisymmetric part
$\Omega_{\alpha\beta}$ are defined as 
\begin{equation}
A_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} + W_{\beta\alpha}),
\end{equation}
and
\begin{equation}
\Omega_{\alpha\beta} = {\textstyle \frac{1}{2}} (W_{\alpha\beta} - W_{\beta\alpha}).
\end{equation}
This full coupling term is then
\begin{eqnarray}
\label{coupling-term}
S_{\alpha\beta}  &=&
(\xi A_{\alpha\nu} + \Omega_{\alpha\nu})
(Q_{\nu\beta} + {\textstyle \frac{1}{3}}\delta_{\nu\beta})
+
(Q_{\alpha\nu} + {\textstyle \frac{1}{3}} \delta_{\alpha\nu})
(\xi A_{\nu\beta} - \Omega_{\nu\beta})
\nonumber\\ 
&-&2 \xi ({Q_{\alpha\beta} + {\textstyle \frac{1}{3}}\delta_{\alpha\beta}})
Q_{\nu\nu} W_{\nu\nu}.
\end{eqnarray}
%CORRECT TRACE?!
Here, $\xi$ is a material-dependent that controls the
relative importance of rotational and elongational flow for molecular
alignment, and determines in practice how the orientational order
responds to a local shear flow. 
In all cases, the value used here is $\xi = 0.7$, which is within
the `flow aligning' regime, where molecules align at a fixed angle
(the Leslie angle) to the flow direction in weak simple shear \cite{deGennes}.
The value of the collective rotational diffusion used in all simulations is
$\Gamma = 0.5$ in simulation units.

@@

In all the simulations reported here the preferred orientation of the
director field at the colloid surface is normal to the surface. The
required nematic director at the particle surface in a given location
$\hat{n}^0_\alpha$ may then be determined from geometry alone,
and a preferred order parameter $Q_{\alpha\beta}^0$ is computed via
\begin{equation}
Q^0_{\alpha\beta} = q^0(\hat{n}_\alpha^0 \hat{n}_\beta^0 
- {\textstyle \frac{1}{3}} \delta_{\alpha\beta}).
\label{eq:surf_q}
\end{equation}
The magnitude of surface order $q^0$ is set by
\begin{equation}
q^0 = {\textstyle \frac{2}{3}} \left( {\textstyle \frac{1}{4}} 
+ {\textstyle \frac{3}{4}} \sqrt{1 + (8/3\gamma)} \right)
\label{eq:surf_q0}
\end{equation}
which corresponds to the bulk order in a defect-free nematically ordered 
sample~\cite{denniston} (this is also very close to the value of the order 
parameter in a cholesteric or in a blue phase away from disclinations).

In all simulations the colloidal (hard sphere) radius is $a = 2.3$. To
prevent colloids overlapping at the hard sphere radius, an additional
short range soft potential is included. This takes the form of
$V(h) = \epsilon (\sigma/h)$ where $h$ is the surface to surface separation.
The parameters are $\epsilon = 0.0004$ and $\sigma = 0.1$ in simulation
units. Both potential and resulting force are smoothly matched to zero at
a cut-off distance of $h_c = 0.25$ simulation units.

@@

To get from simulation to physical units, we need to calibrate scales of length,
energy and time. For similar mappings, see e.g.~\cite{denniston2}.

The length scale can be set by mapping our half pitch to realistic blue phase
unit cell size, which are in the 100-500 nm range \cite{blue1}. 
For bulk simulations (Fig. 1 and Fig. 3 in the main text), a suitable choice
is that that one simulation unit (lattice site) corresponds to 10 nm. Therefore
the colloidal size in Figs. 1 and 3 corresponds to $\sim$ 50 nm.

To get an energy scale, we choose $A_0 \simeq 10^6$ Pa, which is 
reasonable from Ref.~\cite{blue1} (this is also the choice in
Ref.~\cite{oliver2}). This leads to a simulation unit of
free energy (or stress) equal to $10^{8}$Pa in Figs. 1 and 3.
From the energy and length scales one can map elastic constants to
physical units: for instance the simulation of Fig. 1 corresponds to
a liquid crystal with splay, bend and twist (Frank) elastic constants
equal to 35 pN (or equivalently $K=$70 pN).
\begin{equation}
\frac{a}{b^2} = \frac{27}{2 A_0\gamma} \sim 2-5 \times 10^{-6}
{\rm J}^{-1}\, {\rm m}^{3}.
\end{equation}

For the timescale calibration we use the formula (see e.g.~\cite{denniston1})
\begin{equation}
\gamma_1=\frac{2q^2}{\Gamma}\;.
\end{equation}
 which relates the `rotational viscosity' $\gamma_1$, to the ordering strength $q$ and the order parameter mobility $\Gamma$.
In our simulations, we chose $\Gamma = 0.5$, while the value of $\gamma$ used in Fig. 1 leads to $q=1/2$.   
For real liquid crystalline materials, $\gamma_1$ lies usually in range $10^{-2}-1$ Pa s~\cite{deGennes}; we choose for definiteness $\gamma_1 = 1$ Pa s (equivalently 10 poise). Given the previous mapping for free energy (or stress) units, this leads to a simulation unit of time equal to $10^{-8}$ s. The total simulated time in Figs. 1-3 is therefore in the ms range.

To calibrate the electric field strength 
we equate the value of the dimensionless parameter ${\cal E}$ in our 
simulations with that of a hypothetical experiment.
Given the previous value for $A_0$, and a dielectric anisotropy of 10
(or equivalently $\epsilon_a=10\epsilon_0$, with $\epsilon_0$ the
dielectric permittivity of vacuum), we then obtain that an electric field
of 5 V/$\mu$m corresponds to a value of ${\cal E}\sim$0.01.

Given these mappings, all quantities can be easily transformed from
simulation to physical units and vice versa. For example Fig. 1 and 2
use an isotropic viscosity $\eta \simeq$ 0.01 and 0.1 in simulation units,
which translate into $0.01-0.1$ Pa s respectively. This is sensible (if 
slightly low) for a molecular nematogen in the isotropic phase. 
The effective viscosity in ordered phases is higher, but this is ensured
in our model by the coupling to the order parameter.
We note that the density is set to 1 in our simulations: this
corresponds to a fluid density which is about a thousand times larger
than in experiments. This however makes no difference in practice as
the Reynolds number $= \rho V\Lambda/\eta$ (where $V$ is a typical
velocity, in our simulations around $10^{-5}$ in simulation units) 
remains small enough~\cite{codef}.

To summarize the above, our simulations represent BP-forming materials, with the interpretation of the LB units for length, time, and energy density are close to 10 nm, 10 ns, and 100 MPa respectively.

%ISSUE: Temperature?  Temperature: $T = 2.133333 10^{-9}$ CHECK

%ISSUE: no redshift?

%ISSUE: do we want some free energy numbers?

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.35]{s1.jpg}
\end{center}
\caption{As for main manuscript Fig.~1A-E (initialised to equilibrium blue
phase I), but for the entire simulation system of 128$^3$ lattice sites. (a) 
solid volume fraction 1\% and weak anchoring; 
(b) 1\% solid volume fraction and strong anchoring;
(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
volume fraction and strong anchoring. The view direction in the
main figure is from the left here. There is a reference
particle at the bottom left in each panel which does not take part
in the simulation.}
\end{figure}

\begin{figure}
\includegraphics[scale=0.35]{s2.jpg}
\caption{As for main manuscript Fig.~1E-H (initialised via a ``quench''
to generate a disordered network), but for the entire simulation
system of 128$^3$ lattice sites. (a) solid volume fraction 1\% and
weak anchoring; (b) 1\% solid volume fraction and strong anchoring;
(c) 5\% solid volume fraction and weak anchoring; (d) 5\% solid
volume fraction and strong anchoring. Again, there is a reference
particle at the bottom left in each case which does not take part
in the simulation.}
\end{figure}

@@

\bibitem{denniston}
C. Denniston, E. Orlandini, J. M. Yeomans, 
\textit{Phys. Rev. E} \textbf{63}, 056702 (2001).

\bibitem{denniston2}
C. Denniston, D. Marenduzzo, E. Orlandini, J. M.  Yeomans, 
\textit{Philos. Trans. R. Soc. London, Ser. A} \textbf{362}, 1745 (2004).

@@

\bibitem{oliver2}
O. Henrich, K. Stratford, D. Marenduzzo, M. E. Cates, 
\textit{Proc. Acad. Sci. USA} \textbf{107}, 1312 (2010).
